---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.path = "man/figures/README-",
    out.width = "100%"
)
```

# misha

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/misha)](https://CRAN.R-project.org/package=misha)
<!-- badges: end -->

The `misha` package is a toolkit for analysis of genomic data. it implements an efficient data structure for storing genomic data, and provides a set of functions for data extraction, manipulation and analysis.

## Installation

You can install the released version of misha from [CRAN](https://CRAN.R-project.org) with:

```{r, eval=FALSE}
install.packages("misha")
```

And the development version from GitHub with:

```{r, eval=FALSE}
remotes::install_github("tanaylab/misha")
```

## Usage 

See the [Genomes](https://tanaylab.github.io/misha/articles/Genomes.html) vignette for instructions on how to create a misha database for common genomes.

See the [user manual](https://tanaylab.github.io/misha/articles/Manual.html) for more usage details.







