---
title: "Genomes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Genomes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

```{r setup}
library(misha)
```

# Create a `misha` database from UCSC

The easiest way to create a `misha` database is to use the `gdb.create_genome` function:

```{r, eval = FALSE}
gdb.create_genome("hg19") # creates a database for the hg19 genome
gdb.create_genome("hg38") # creates a database for the hg38 genome
gdb.create_genome("mm10") # creates a database for the mm10 genome
gdb.create_genome("mm9") # creates a database for the mm9 genome
gdb.create_genome("mm39") # creates a database for the mm39 genome
```

However, if you need to create a database for a genome that is not supported by `gdb.create_genome`, or if you want to make sure that the database is created from the latest version of the genome in ucsc, you can create it manually using the commands below. 

## hg19
In order to create a misha database for _hg19_ genome, run the following commands (assuming *"hg19"* is your new data base path):

```{r, eval = FALSE}
ftp <- "ftp://hgdownload.soe.ucsc.edu/goldenPath/hg19"
gdb.create(
    "hg19",
    paste(ftp, "chromosomes", paste0("chr", c(1:22, "X", "Y", "M"), ".fa.gz"), sep = "/"),
    paste(ftp, "database/knownGene.txt.gz", sep = "/"),
    paste(ftp, "database/kgXref.txt.gz", sep = "/"),
    c(
        "kgID", "mRNA", "spID", "spDisplayID", "geneSymbol",
        "refseq", "protAcc", "description", "rfamAcc",
        "tRnaName"
    )
)
gdb.init("hg19")
```

## hg38
In order to create a misha database for _hg38_ genome, run the following commands (assuming *"hg38"* is your new data base path):

```{r, eval = FALSE}
ftp <- "ftp://hgdownload.soe.ucsc.edu/goldenPath/hg38"
gdb.create(
    "hg38",
    paste(ftp, "chromosomes", paste0("chr", c(1:22, "X", "Y", "M"), ".fa.gz"), sep = "/"),
    paste(ftp, "database/knownGene.txt.gz", sep = "/"),
    paste(ftp, "database/kgXref.txt.gz", sep = "/"),
    c(
        "kgID", "mRNA", "spID", "spDisplayID", "geneSymbol",
        "refseq", "protAcc", "description", "rfamAcc",
        "tRnaName"
    )
)
gdb.init("hg38")
```

## mm9
In order to create a misha database for _mm9_ genome, run the following commands (assuming *"mm9"* is your new data base path):

```{r, eval = FALSE}
ftp <- "ftp://hgdownload.soe.ucsc.edu/goldenPath/mm9"
gdb.create(
    "mm9",
    paste(ftp, "chromosomes", paste0("chr", c(1:19, "X", "Y", "M"), ".fa.gz"), sep = "/"),
    paste(ftp, "database/knownGene.txt.gz", sep = "/"),
    paste(ftp, "database/kgXref.txt.gz", sep = "/"),
    c(
        "kgID", "mRNA", "spID", "spDisplayID", "geneSymbol",
        "refseq", "protAcc", "description"
    )
)
gdb.init("mm9")
```

## mm10
In order to create a misha database for _mm10_ genome, run the following commands (assuming *"mm10"* is your new data base path):

```{r, eval = FALSE}
ftp <- "ftp://hgdownload.soe.ucsc.edu/goldenPath/mm10"
gdb.create(
    "mm10",
    paste(ftp, "chromosomes", paste0("chr", c(1:19, "X", "Y", "M"), ".fa.gz"), sep = "/"),
    paste(ftp, "database/knownGene.txt.gz", sep = "/"),
    paste(ftp, "database/kgXref.txt.gz", sep = "/"),
    c(
        "kgID", "mRNA", "spID", "spDisplayID", "geneSymbol",
        "refseq", "protAcc", "description", "rfamAcc",
        "tRnaName"
    )
)
gdb.init("mm10")
```

## mm39

In order to create a misha database for _mm39_ genome, run the following commands (assuming *"mm39"* is your new data base path):

```{r, eval = FALSE}
ftp <- "ftp://hgdownload.soe.ucsc.edu/goldenPath/mm39"
gdb.create(
    "mm39",
    paste(ftp, "chromosomes", paste0("chr", c(1:19, "X", "Y", "M"), ".fa.gz"), sep = "/"),
    paste(ftp, "database/knownGene.txt.gz", sep = "/"),
    paste(ftp, "database/kgXref.txt.gz", sep = "/"),
    c(
        "kgID", "mRNA", "spID", "spDisplayID", "geneSymbol",
        "refseq", "protAcc", "description", "rfamAcc",
        "tRnaName"
    )
)
gdb.init("mm39")
```

## Building from UCSC mammal hubs (Zoonomia)

UCSC mammal hubs host 224 of the 241 Cactus / Zoonomia mammalian genomes
under a uniform per-species directory layout, including chromAlias, repeat
masker output, and one or more GTFs per species. Build a misha groot from a
hub assembly with:

```{r, eval = FALSE}
# One-shot. Default sets = c("genes", "rmsk", "cgi", "cytoband"); cytoband and
# (often) cgi are skipped with a warning since hubs don't ship them.
gdb.build_genome("GCA_004023825.1",
    path   = "/genomes/arctic_fox",
    prefix = "intervs.global."
)
```

If you already have a private FASTA build for the same accession (or a HAL-
derived assembly) and just want to layer canonical annotations on it:

```{r, eval = FALSE}
gdb.install_intervals(
    groot  = "/genomes/my_private_assembly",
    source = list(source = "ucsc-hub", accession = "GCA_004023825.1"),
    sets   = c("genes", "rmsk"),
    prefix = "intervs.global."
)
```

If your groot was built with a different chromosome-naming scheme than the
hub uses (for example, Cactus-derived names), the hub's `chromAlias.txt` is
consulted to translate annotation chroms into the groot's canonical names —
strictly. The chosen alias column must contain *every* groot chrom and *every*
distinct chrom in the source file; otherwise the install errors with a
diagnostic per-column overlap table.

To convert hub GTFs the install needs UCSC's `gtfToGenePred` binary; on the
first call it will prompt for consent to download (~5 MB). To pre-install in
non-interactive contexts:

```{r, eval = FALSE}
gdb.install_gtf_converter()
# Or set MISHA_GTF_TO_GENEPRED to a binary you provide (e.g. via conda).
```
