---
title: "PWM Virtual Tracks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PWM Virtual Tracks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 7, fig.height = 4
)
```

This vignette covers misha's PWM virtual track family: the existing
aggregating scorers (`pwm`, `pwm.max`, `pwm.max.pos`, `pwm.count`), the
edit-distance variants, and the per-position attribution tracks
(`pwm.grad`, `pwm.grad.ism`).

The aggregating sections (`pwm`, `pwm.max`, ...) are well documented on
the `gvtrack.create` help page, so the focus here is on the mental
model and on examples a reader can adapt to their own database. The
math for the gradient tracks is worked out in more detail because they
are new and there is no existing reference.

The math chunks below run as plain R (no genome database required) so
they always evaluate. The genome-track examples are shown with
`eval = FALSE` because they need a database; the patterns transfer
unchanged to any misha DB.

## What is a PSSM?

A position-specific scoring matrix (PSSM, also "PWM") is a per-column
distribution over A/C/G/T. In misha the input is an `L x 4` matrix:

- **Rows**: motif columns, in 5' -> 3' order (one row per nucleotide
  position in the motif).
- **Columns**: bases, named `A`, `C`, `G`, `T` (any extra columns are
  ignored).

The raw matrix typically holds frequencies or counts. Internally misha
adds a Dirichlet pseudocount (`prior`, default `0.01`), normalizes each
row to sum to 1, and takes the natural log. The result is a matrix of
log-probabilities `M[c, b]`: the log-likelihood that base `b` appears
at column `c` of the motif.

```{r}
# A 3-column PSSM with consensus ACG.
pssm <- matrix(
    c(
        0.97, 0.01, 0.01, 0.01,
        0.01, 0.97, 0.01, 0.01,
        0.01, 0.01, 0.97, 0.01
    ),
    nrow = 3, ncol = 4, byrow = TRUE,
    dimnames = list(NULL, c("A", "C", "G", "T"))
)
pssm
```

```{r}
# Apply prior + per-row normalization, then log.
prior <- 0.01
M <- log((pssm + prior) / rowSums(pssm + prior))
round(M, 3)
```

`M[1, "A"]` is close to 0 (high probability), while every off-consensus
entry is a large negative number. The `prior` keeps log-zero rows
finite: with `prior = 0` and a strict frequency matrix, missing bases
become `-Inf` and dominate any sum. `0.01` is a conservative default; a
raw count matrix often wants something larger.

## Per-anchor score

Given a sequence `s`, an *anchor* is a position `a` at which the motif
is laid down. The per-anchor (single-strand, forward) score is

$$
\text{score}_a \;=\; \sum_{c=0}^{L-1} M[c,\, s_{a+c}].
$$

It is just a column-by-column lookup followed by a sum. Units are log
probability; values are <= 0 by construction (each term is the log of a
probability in [0, 1]).

```{r}
# Score the anchor at position 1 of the sequence "ACGTTT".
seq <- "ACGTTT"
bases <- strsplit(seq, "")[[1]]
score_anchor <- function(M, bases, anchor) {
    L <- nrow(M)
    sum(vapply(seq_len(L), function(c) M[c, bases[anchor + c - 1]], numeric(1)))
}
score_anchor(M, bases, 1) # "ACG" - perfect consensus
score_anchor(M, bases, 2) # "CGT" - off-consensus
```

The first anchor matches the consensus and scores close to 0; the
second is well below.

For a window of length `W >= L`, we get `W - L + 1` per-anchor scores.

```{r}
all_anchor_scores <- function(M, seq) {
    bases <- strsplit(seq, "")[[1]]
    W <- length(bases)
    L <- nrow(M)
    vapply(seq_len(W - L + 1), function(a) score_anchor(M, bases, a), numeric(1))
}
all_anchor_scores(M, seq)
```

## Aggregating over an interval

The PWM virtual tracks each reduce these per-anchor scores to a single
scalar per iterator interval.

### `pwm.max`: best anchor

`pwm.max` returns the maximum per-anchor score in the window. Easiest
to interpret ("how good is the best motif hit here?"), but discards
information about secondary hits.

```{r}
max(all_anchor_scores(M, seq))
```

### `pwm` (LSE): smooth aggregate

`pwm` returns the log-sum-exp of all per-anchor scores:

$$
f_{\text{LSE}} \;=\; \log \sum_a \exp(\text{score}_a).
$$

This is dominated by the best anchor when one anchor clearly wins, but
also adds support for many mediocre anchors (an "OR" over noisy
matches). LSE is also smooth: small changes in the input give small
changes in the output, which is friendlier for downstream filtering and
gradient-based attribution.

```{r}
log_sum_exp <- function(x) {
    m <- max(x)
    m + log(sum(exp(x - m)))
}
s <- all_anchor_scores(M, seq)
log_sum_exp(s)
```

For a single dominant hit `pwm` and `pwm.max` are nearly equal; for a
window with many weak similar-strength hits `pwm` can be substantially
larger.

### `pwm.max.pos`: argmax position

`pwm.max.pos` returns the 1-based offset of the argmax anchor inside
the iterator interval, signed by strand when `bidirect = TRUE`. Useful
when you want to know *where* the best hit sits within a region.

### `pwm.count`: thresholded hit count

`pwm.count` requires `score.thresh`. It counts anchors whose score is
at or above `score.thresh`. With `bidirect = TRUE` the union over
strands is taken at each genomic start, so palindromic hits count once.

```{r, eval = FALSE}
gvtrack.create("motif_count", NULL, "pwm.count",
    pssm = pssm, score.thresh = -2, bidirect = TRUE
)
gvtrack.iterator("motif_count", sshift = 0, eshift = 500)
gextract("motif_count", iterator = 500, intervals = my_intervals)
```

### Why LSE rather than just MAX?

For any aggregation over a long window, LSE has three practical
advantages over MAX:

1. **Integrates many anchors.** A region with five mediocre hits has a
   higher LSE than one with a single mediocre hit, but the same MAX.
2. **Smooth.** Differentiable everywhere; small input changes -> small
   output changes. This matters for gradient attribution
   (next section).
3. **Numerically stable when implemented with the standard
   `m + log(sum(exp(x - m)))` trick.**

When you only care about "is there a strong hit anywhere in this
window?", `pwm.max` is fine. For typical "motif energy" scoring across
500 bp - 1 kb windows, `pwm` (LSE) is the better default.

## Common parameters

These mirror what `?gvtrack.create` documents; one-liner each:

- **`bidirect`** (default `TRUE`): scan both strands and combine per
  genomic start. With `bidirect = FALSE`, only the strand given by
  `strand` is scanned.
- **`strand`** (`+1` / `-1`): only used when `bidirect = FALSE`.
- **`prior`** (default `0.01`): Dirichlet pseudocount added to PSSM
  rows before normalization. Raise it for noisy / count-based PSSMs;
  set to 0 to disable.
- **`extend`** (default `TRUE`): pad the fetched sequence so anchors
  near the right edge of the iterator interval still get a full
  window. Without extension, an interval of length `W` only yields
  `W - L + 1` anchors.
- **`spat_factor`** + **`spat_bin`**: optional position-dependent log
  weights. `spat_factor` is a positive numeric vector; `spat_bin` is
  the bin width in bp. Each anchor's score is multiplied by the log
  factor for its bin. Use this to weight central bases of a window
  more than flanks.
- **`score.thresh`**: required by `pwm.count` and the edit-distance
  variants. Anchors with score >= threshold are "hits". Rejected by
  `pwm.grad` and `pwm.grad.ism`.

## Edit-distance variants

These ask "how many single-base edits would it take to make / break a
motif match?". They share the same PSSM machinery but answer a
discrete combinatorial question instead of returning a continuous
score.

- **`pwm.edit_distance`**: minimum number of edits (substitutions, and
  optionally indels via `max_indels`) needed across all windows in the
  iterator interval to push the score across `score.thresh`.
  `direction = "above"` (default) targets making a match;
  `direction = "below"` targets disrupting one.
- **`pwm.max.edit_distance`**: same idea, but always evaluated at the
  argmax window of `pwm.max`.
- **`pwm.n_mutations`**: number of independent single-base
  substitutions that each cross `score.thresh` on their own. Returns 0
  if the threshold is already met, NA if no single substitution
  suffices.

There is also an LSE-aware family (`pwm.edit_distance.lse`,
`pwm.edit_distance.lse.pos`) that targets the LSE aggregate rather
than per-window scores.

See `?gvtrack.create` for the full parameter list and worked examples.

## Per-position attribution: `pwm.grad` and `pwm.grad.ism`

The aggregating tracks above all collapse a window into one scalar.
`pwm.grad` and `pwm.grad.ism` answer a different question: **at this
specific base pair, how much does the local sequence support the PWM
aggregate?** The output is one non-negative number per iterator
interval, computed at the anchor that starts at `interval.start`.

Combined with a `iterator = 1` step and `gvtrack.iterator(eshift =
W - 1)`, this gives a per-bp saliency / attribution track you can plot
or combine with conservation, ATAC, methylation, etc.

### Motivation

Two use cases:

1. **Saliency.** "Of the base pairs in this window, which ones are
   actually carrying the motif signal?" - a poor man's
   integrated-gradient / DeepLIFT.
2. **Single-bp coupling.** Combine PWM evidence with another per-bp
   signal (conservation, ChIP, etc.) at single-base resolution. With
   `pwm` you only get one number per window; with `pwm.grad` you get
   one per bp.

### Math intuition (linearized, `pwm.grad`)

Think of each base as a one-hot vector `x_p` over A/C/G/T. The score
`score_a` is linear in `x_p` whenever `a` contains `p`, with gradient
`M[p - a, b]`. With pivot fixed at the iterator start `p =
interval.start`, only one anchor (`a = p`) contains `p`, and only
column 0 of the PSSM is involved.

For the LSE aggregate, the gradient is the softmax-weighted column-0
score, contracted against `(x_p - x_p^{\text{worst}})` where the
reference is the worst possible base at column 0:

$$
g_{\text{LSE}}(p) \;=\; w_p \cdot \big(M[0,\, b_p] - \min_b M[0,\, b]\big),
\qquad w_p = \exp(\text{score}_p - f_{\text{LSE}}).
$$

Three useful sanity points:

- **Non-negative.** `M[0, b_p] - min_b M[0, b]` >= 0 by construction;
  `w_p` in [0, 1].
- **Bounded by per-column information.** The diff term equals the
  column's information content under the "actual minus worst" baseline.
- **Modulated by softmax weight.** If the head anchor dominates the
  LSE (`w_p ≈ 1`), the gradient saturates to the diff. If the LSE
  spreads over many anchors (`w_p` small), the gradient is small even
  if the head anchor is itself a good match - the *attribution* of
  this particular base to the aggregate is diluted.

For the MAX aggregate, the gradient picks up the diff iff the head
anchor is the argmax:

$$
g_{\text{MAX}}(p) \;=\;
\begin{cases}
M[0,\, b_p] - \min_b M[0,\, b] & \text{if } a^* = p,\\
0 & \text{otherwise.}
\end{cases}
$$

### Math intuition (in-silico mutagenesis, `pwm.grad.ism`)

The ISM gradient is "actual minus best replacement":

$$
g_{\text{ISM}}(p) \;=\; f(\text{seq}) \;-\; \min_{b' \neq b_p} f\big(\text{seq with } s_p := b'\big).
$$

Always >= 0. With pivot fixed at `interval.start`, only `score_p`
changes when we flip `s_p`, so the update is O(1) per flip and ISM
costs only ~3-4x more than the linearized form.

### When `pwm.grad` and `pwm.grad.ism` differ

The two definitions agree exactly when the argmax is stable under
flips of `s_p`, and agree to high precision when LSE concentrates on
the head anchor. They diverge in two cases:

1. **Argmax shifts.** For MAX, if flipping `s_p` makes a *different*
   anchor become the argmax, ISM accounts for that ("the second-best
   hit kicks in") and the linearized form doesn't.
2. **Curvature in LSE.** Linearization is a first-order approximation
   around the actual sequence; ISM evaluates the aggregate at three
   discrete neighbours and takes the worst. For sharply non-linear
   regions of the LSE landscape they differ slightly, but in practice
   the disagreement is small.

In short: **`pwm.grad` is cheaper and almost always good enough;
`pwm.grad.ism` is more conservative around argmax-shift edge cases.**

### Worked numerical example

Consider an `L = 3` PSSM and a 6-bp window starting with consensus
"ACG":

```{r}
pssm <- matrix(
    c(
        0.97, 0.01, 0.01, 0.01,
        0.01, 0.97, 0.01, 0.01,
        0.01, 0.01, 0.97, 0.01
    ),
    nrow = 3, ncol = 4, byrow = TRUE,
    dimnames = list(NULL, c("A", "C", "G", "T"))
)
M <- log((pssm + 0.01) / rowSums(pssm + 0.01))

seq <- "ACGTTT"
s <- all_anchor_scores(M, seq) # 4 per-anchor scores
s
```

Linearized LSE gradient at `p = 1`:

```{r}
f_lse <- log_sum_exp(s)
score_p <- s[1]
w_p <- exp(score_p - f_lse) # softmax weight at the head
b_p <- "A" # head base
diff_col0 <- M[1, b_p] - min(M[1, ])
w_p
diff_col0
g_lse <- w_p * diff_col0
g_lse
```

`w_p ≈ 1` because the head anchor "ACG" is the dominant LSE
contributor; `diff_col0` is the column-0 information content. The
result is essentially the column's entropy gap.

ISM gradient: flip `seq[1]` to each of C/G/T and recompute the LSE.

```{r}
ism_lse <- function(M, seq) {
    s <- all_anchor_scores(M, seq)
    log_sum_exp(s)
}
flips <- setdiff(c("A", "C", "G", "T"), substr(seq, 1, 1))
f_alt <- vapply(flips, function(b) {
    ism_lse(M, paste0(b, substr(seq, 2, nchar(seq))))
}, numeric(1))
f_alt
ism_lse(M, seq) - min(f_alt)
```

The two values agree to within numerical noise; this is the typical
case.

### Genome-track recipe

Here is the canonical pattern: extract `pwm` and `pwm.grad` over a 1
kb iterator window, then walk the per-bp gradient with `iterator = 1`.

```{r, eval = FALSE}
W <- 500 # window length (bp) used for both aggregate and gradient

pssm <- matrix(...) # your 4-col PSSM, dimnames A/C/G/T

gvtrack.create("score", NULL, "pwm", pssm = pssm)
gvtrack.create("grad", NULL, "pwm.grad", pssm = pssm, aggregate = "lse")
gvtrack.create("grad_ism", NULL, "pwm.grad.ism", pssm = pssm, aggregate = "lse")

# Each per-bp value uses a window starting at this bp and extending W bp right.
gvtrack.iterator("score", sshift = 0, eshift = W - 1)
gvtrack.iterator("grad", sshift = 0, eshift = W - 1)
gvtrack.iterator("grad_ism", sshift = 0, eshift = W - 1)

regions <- gintervals(1, 100000, 110000)
out <- gextract(c("score", "grad", "grad_ism"),
    intervals = regions, iterator = 1
)
head(out)
```

Common downstream patterns:

- **Per-bp coupling**: `out$grad * out$other_track` to get an
  attribution-weighted version of `other_track`.
- **Argmax-shift hunting**: `out$grad_ism - out$grad` highlights bp
  where ISM and linearization diverge, i.e. where the argmax flips
  under a single substitution.
- **Smoothing**: a `gtrack.smooth` over `pwm.grad` gives a soft
  saliency band over the canonical motif location.

### `bidirect = TRUE` for asymmetric PSSMs

For palindromic PSSMs the rc strand contributes column-0 of the rc
PSSM (= column `L-1` of the fwd PSSM, with complement). With
`bidirect = TRUE` the gradient uses a per-anchor strand softmax to
combine the two contributions; this is handled internally and you
don't need to touch anything beyond `bidirect = TRUE`. For asymmetric
PSSMs, the bidirect gradient at a bp can pick up signal from either
strand depending on which one wins locally.

## Performance notes

In rough numbers for an `L = 12` PSSM, `W = 500` bp window, scanned at
1 bp resolution:

- `pwm.grad` is roughly comparable to `pwm` (both do the same scan;
  the gradient post-step is O(1) per anchor).
- `pwm.grad.ism` is ~3-4x `pwm.grad` (three flips x O(1) per
  position, with a small constant for log-space arithmetic).

For genome-wide attribution, you almost never want `iterator = 1`
across the whole genome. Common patterns:

- Use a sparse intervals set as iterator (e.g. peaks, TSS windows) and
  `iterator = 1` only inside those; gives single-bp attribution where
  it matters and skips empty space.
- Use a coarser iterator step (e.g. `iterator = 10` or `iterator =
  100`) for whole-genome scans; gradient is still meaningful, just
  averaged over the step.

When in doubt, profile on one chromosome before committing to a
genome-wide scan: the expensive part is reading sequence from disk,
not the gradient computation itself.
