---
title: "PWM Virtual Tracks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PWM Virtual Tracks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 7, fig.height = 4
)
```

This vignette covers misha's PWM virtual track family: the aggregating
scorers (`pwm`, `pwm.max`, `pwm.max.pos`, `pwm.count`), the
edit-distance variants, and the per-position attribution tracks
(`pwm.grad`, `pwm.grad.ism`).

The math chunks below run as plain R (no genome database required) so
they always evaluate. Genome-track examples are shown with
`eval = FALSE` because they need a database; the patterns transfer
unchanged to any misha DB.

## What is a PSSM?

A position-specific scoring matrix (PSSM, also "PWM") is a per-column
distribution over A/C/G/T. In misha the input is an `L x 4` matrix:

- **Rows**: motif columns, in 5' -> 3' order (one row per nucleotide
  position in the motif).
- **Columns**: bases, named `A`, `C`, `G`, `T` (any extra columns are
  ignored).

The raw matrix typically holds frequencies or counts. Internally misha
adds a Dirichlet pseudocount (`prior`, default `0.01`), normalizes each
row to sum to 1, and takes the natural log. The result is a matrix of
log-probabilities `M[c, b]`: the log-likelihood that base `b` appears
at column `c` of the motif.

```{r}
# A 3-column PSSM with consensus ACG.
pssm <- matrix(
    c(
        0.97, 0.01, 0.01, 0.01,
        0.01, 0.97, 0.01, 0.01,
        0.01, 0.01, 0.97, 0.01
    ),
    nrow = 3, ncol = 4, byrow = TRUE,
    dimnames = list(NULL, c("A", "C", "G", "T"))
)
pssm
```

```{r}
# Apply prior + per-row normalization, then log.
prior <- 0.01
M <- log((pssm + prior) / rowSums(pssm + prior))
round(M, 3)
```

`M[1, "A"]` is close to 0 (high probability), while every off-consensus
entry is a large negative number. The `prior` keeps log-zero rows
finite: with `prior = 0` and a strict frequency matrix, missing bases
become `-Inf` and dominate any sum. `0.01` is a conservative default; a
raw count matrix often wants something larger.

## Per-anchor score

Given a sequence `s`, an *anchor* is a position `a` at which the motif
is laid down. The forward-strand per-anchor score is

$$
\text{score}_a^{\text{fwd}} \;=\; \sum_{c=0}^{L-1} M[c,\, s_{a+c}].
$$

It is just a column-by-column lookup followed by a sum. Units are log
probability; values are <= 0 by construction.

```{r}
# Score the anchor at position 1 of the sequence "ACGTTT".
seq <- "ACGTTT"
bases <- strsplit(seq, "")[[1]]
score_anchor <- function(M, bases, anchor) {
    L <- nrow(M)
    sum(vapply(seq_len(L), function(c) M[c, bases[anchor + c - 1]], numeric(1)))
}
score_anchor(M, bases, 1) # "ACG" - perfect consensus
score_anchor(M, bases, 2) # "CGT" - off-consensus
```

For a window of length `W >= L`, we get `W - L + 1` per-anchor scores.

```{r}
all_anchor_scores <- function(M, seq) {
    bases <- strsplit(seq, "")[[1]]
    W <- length(bases)
    L <- nrow(M)
    vapply(seq_len(W - L + 1), function(a) score_anchor(M, bases, a), numeric(1))
}
all_anchor_scores(M, seq)
```

## Strands and `bidirect`

Anchors are always indexed in **forward-genome coordinates**: anchor
`a` covers genomic positions `[a, a + L)`. What changes with the
strand setting is the *score* attached to each anchor.

For each genomic anchor `a` there are two scores:

- **Forward score** `score_a^{fwd}`: PSSM `M` applied directly to the
  bases at `[a, a + L)`. Column `c` of the PSSM looks up base `s_{a+c}`.
- **Reverse score** `score_a^{rc}`: PSSM applied to the
  reverse-complement of `[a, a + L)`. Equivalently, column `c` of the
  PSSM looks up the *complement* of `s_{a + L - 1 - c}`.

`bidirect` and `strand` decide which of these the per-anchor aggregate
uses:

| Setting | Per-anchor score |
|---|---|
| `bidirect = TRUE` (default) | `log(exp(score_a^{fwd}) + exp(score_a^{rc}))` (per-genomic-position union over strands) |
| `bidirect = FALSE, strand = 1` | `score_a^{fwd}` |
| `bidirect = FALSE, strand = -1` | `score_a^{rc}` |

The `bidirect = TRUE` case is the most common. The combined score is
the LSE union of the two strands per anchor; equivalent to "what's the
best motif support at this genomic position, considering both
orientations". This is computed at *every* anchor, then aggregated
(MAX / LSE / COUNT) over the iterator window.

The `strand` argument is **only** used when `bidirect = FALSE`.

### Position reporting

Several tracks return positions:

- **`pwm.max.pos`**: 1-based offset of the argmax anchor inside the
  iterator interval (after `gvtrack.iterator()` shifts). With
  `bidirect = TRUE`, the value is **signed**: positive when the
  forward strand wins at the argmax, negative when the reverse strand
  wins. The absolute value is always the offset, never a genomic
  coordinate.
- **`pwm.grad`, `pwm.grad.ism`**: report the gradient at a single
  position - the iterator interval **start**, in forward-genome
  coordinates, regardless of strand.

So if you want a per-bp gradient track at every position `p` in a
region, you call `gextract` with `iterator = 1` and an iterator
extension `gvtrack.iterator(eshift = W - 1)` so each per-bp evaluation
sees a window `[p, p + W)`. The reported value is at `p`, in
fwd-genome coords, exactly the position you iterated to.

## Aggregating over an interval

The aggregating PWM virtual tracks reduce the per-anchor scores in an
iterator window to a single scalar.

### `pwm.max`: best anchor

`pwm.max` returns the maximum per-anchor score in the window.

```{r}
max(all_anchor_scores(M, seq))
```

### `pwm` (LSE): integrated likelihood

`pwm` returns the log-sum-exp of all per-anchor scores:

$$
f_{\text{LSE}} \;=\; \log \sum_a \exp(\text{score}_a).
$$

This is dominated by the best anchor when one anchor clearly wins, but
also picks up support from many mediocre anchors (an "OR" over noisy
matches). Useful when you care about cumulative motif evidence in a
region rather than just the single best hit.

```{r}
log_sum_exp <- function(x) {
    m <- max(x)
    m + log(sum(exp(x - m)))
}
s <- all_anchor_scores(M, seq)
log_sum_exp(s)
```

For a single dominant hit `pwm` and `pwm.max` are nearly equal; for a
window with many similar-strength hits `pwm` can be substantially
larger.

### `pwm.max.pos`: argmax position

Returns the offset of the argmax anchor inside the iterator interval.
See "Position reporting" above for the sign convention.

### `pwm.count`: thresholded hit count

`pwm.count` requires `score.thresh`. It counts anchors whose score is
at or above `score.thresh`. With `bidirect = TRUE` the per-anchor
score is the strand union, so palindromic hits count once.

```{r, eval = FALSE}
gvtrack.create("motif_count", NULL, "pwm.count",
    pssm = pssm, score.thresh = -2, bidirect = TRUE
)
gvtrack.iterator("motif_count", sshift = 0, eshift = 500)
gextract("motif_count", iterator = 500, intervals = my_intervals)
```

## Common parameters

These mirror what `?gvtrack.create` documents; one-liner each:

- **`bidirect`** (default `TRUE`): see "Strands and `bidirect`" above.
- **`strand`** (`+1` / `-1`): only used when `bidirect = FALSE`.
- **`prior`** (default `0.01`): Dirichlet pseudocount added to PSSM
  rows before normalization. Raise it for noisy / count-based PSSMs;
  set to 0 to disable.
- **`extend`** (default `TRUE`): pad the fetched sequence so anchors
  near the right edge of the iterator interval still get a full
  window. Without extension, an interval of length `W` only yields
  `W - L + 1` anchors.
- **`spat_factor`** + **`spat_bin`**: optional position-dependent log
  weights. `spat_factor` is a positive numeric vector; `spat_bin` is
  the bin width in bp. Each anchor's score is multiplied by the log
  factor for its bin. Use this to weight central bases of a window
  more than flanks.
- **`score.thresh`**: required by `pwm.count` and the edit-distance
  variants. Anchors with score >= threshold are "hits". Rejected by
  `pwm.grad` and `pwm.grad.ism`.

## Edit-distance variants

These ask "how many single-base edits would it take to make / break a
motif match?". They share the same PSSM machinery but answer a
discrete combinatorial question instead of returning a continuous
score.

- **`pwm.edit_distance`**: minimum number of edits (substitutions, and
  optionally indels via `max_indels`) needed across all windows in the
  iterator interval to push the score across `score.thresh`.
  `direction = "above"` (default) targets making a match;
  `direction = "below"` targets disrupting one.
- **`pwm.max.edit_distance`**: same idea, but always evaluated at the
  argmax window of `pwm.max`.
- **`pwm.n_mutations`**: number of independent single-base
  substitutions that each cross `score.thresh` on their own. Returns 0
  if the threshold is already met, NA if no single substitution
  suffices.

There is also an LSE-aware family (`pwm.edit_distance.lse`,
`pwm.edit_distance.lse.pos`) that targets the LSE aggregate rather
than per-window scores.

See `?gvtrack.create` for the full parameter list and worked examples.

## Per-position attribution: `pwm.grad` and `pwm.grad.ism`

The aggregating tracks above collapse a window into one scalar.
`pwm.grad` and `pwm.grad.ism` return one scalar per iterator interval
**evaluated at a specific bp** - the anchor that starts at
`interval.start`. With `iterator = 1` and `gvtrack.iterator(eshift =
W - 1)`, this gives a per-bp track over a region.

### Linearized gradient (`pwm.grad`)

Treat each base as a one-hot vector `x_p` over A/C/G/T. With pivot
fixed at `p = interval.start`, only one anchor (`a = p`) contains `p`,
and only column 0 of the PSSM is involved.

For the LSE aggregate:

$$
g_{\text{LSE}}(p) \;=\; w_p \cdot \big(M[0,\, b_p] - \min_b M[0,\, b]\big),
\qquad w_p = \exp(\text{score}_p - f_{\text{LSE}}).
$$

`w_p` is the softmax weight of the head anchor in the window: it
captures how much the head anchor contributes to the LSE relative to
the other `W - L` anchors. The diff term equals the column-0
information content under an "actual minus worst" baseline.

For the MAX aggregate:

$$
g_{\text{MAX}}(p) \;=\;
\begin{cases}
M[0,\, b_p] - \min_b M[0,\, b] & \text{if } a^* = p,\\
0 & \text{otherwise.}
\end{cases}
$$

i.e. the column-0 diff iff the head anchor is the argmax of the window.

### In-silico mutagenesis (`pwm.grad.ism`)

$$
g_{\text{ISM}}(p) \;=\; f(\text{seq}) \;-\; \min_{b' \neq b_p} f\big(\text{seq with } s_p := b'\big).
$$

Flip the head base to each of the three other letters, recompute the
aggregate, take the worst, and subtract from the unperturbed
aggregate. With pivot at `interval.start`, only `score_p` changes
under the flip, so the update is O(1) per alternative.

### When the two definitions differ

The two agree exactly when the argmax is stable under flips of `s_p`,
and agree to high precision when LSE concentrates on the head anchor.
They diverge when:

1. **Argmax shifts under flip.** For MAX, flipping `s_p` may make a
   *different* anchor become the argmax; ISM picks that up and the
   linearized form doesn't.
2. **LSE curvature.** Linearization is first-order around the actual
   sequence; ISM evaluates the aggregate at three discrete neighbours.
   The disagreement is small in practice.

`pwm.grad` is cheaper; `pwm.grad.ism` is more conservative around
argmax-shift cases.

### Strands

The pivot is at `interval.start` in fwd-genome coordinates regardless
of strand. What changes is which PSSM column matters:

- `bidirect = FALSE, strand = 1`: column 0 of `M`, base `s_p`.
- `bidirect = FALSE, strand = -1`: column `L-1` of `M`, complement of
  `s_p` (because the rc-strand PSSM applied at `p` reads its
  *first* column from `s_p` complemented, which corresponds to fwd
  PSSM column `L-1`).
- `bidirect = TRUE`: per-strand softmax at the head anchor combines
  the two contributions. You don't need to think about it explicitly;
  pass `bidirect = TRUE` and the engine computes the appropriate
  combined gradient.

### Worked numerical example

```{r}
pssm <- matrix(
    c(
        0.97, 0.01, 0.01, 0.01,
        0.01, 0.97, 0.01, 0.01,
        0.01, 0.01, 0.97, 0.01
    ),
    nrow = 3, ncol = 4, byrow = TRUE,
    dimnames = list(NULL, c("A", "C", "G", "T"))
)
M <- log((pssm + 0.01) / rowSums(pssm + 0.01))

seq <- "ACGTTT"
s <- all_anchor_scores(M, seq)
s
```

Linearized LSE gradient at `p = 1`:

```{r}
f_lse <- log_sum_exp(s)
score_p <- s[1]
w_p <- exp(score_p - f_lse)
b_p <- "A"
diff_col0 <- M[1, b_p] - min(M[1, ])
g_lse <- w_p * diff_col0
c(w_p = w_p, diff_col0 = diff_col0, g_lse = g_lse)
```

ISM gradient: flip `seq[1]` to each of C/G/T and recompute the LSE.

```{r}
ism_lse <- function(M, seq) log_sum_exp(all_anchor_scores(M, seq))
flips <- setdiff(c("A", "C", "G", "T"), substr(seq, 1, 1))
f_alt <- vapply(flips, function(b) {
    ism_lse(M, paste0(b, substr(seq, 2, nchar(seq))))
}, numeric(1))
ism_lse(M, seq) - min(f_alt)
```

The two values agree to within numerical noise; this is the typical
case.

### Genome-track recipe

```{r, eval = FALSE}
W <- 500 # window length

pssm <- matrix(...) # your 4-col PSSM, dimnames A/C/G/T

gvtrack.create("score", NULL, "pwm", pssm = pssm)
gvtrack.create("grad", NULL, "pwm.grad", pssm = pssm, aggregate = "lse")
gvtrack.create("grad_ism", NULL, "pwm.grad.ism", pssm = pssm, aggregate = "lse")

gvtrack.iterator("score", sshift = 0, eshift = W - 1)
gvtrack.iterator("grad", sshift = 0, eshift = W - 1)
gvtrack.iterator("grad_ism", sshift = 0, eshift = W - 1)

regions <- gintervals(1, 100000, 110000)
out <- gextract(c("score", "grad", "grad_ism"),
    intervals = regions, iterator = 1
)
head(out)
```

`out$grad - out$grad_ism` flags positions where the argmax flips under
single substitution. Multiplying `out$grad` by another per-bp track
gives an attribution-weighted version of that track.

## Performance notes

For an `L = 12` PSSM, `W = 500` bp window, scanned at 1 bp resolution:

- `pwm` and `pwm.max` use a sliding-window cache and are very fast
  (O(L) per pivot amortized).
- `pwm.grad` and `pwm.grad.ism` use a per-anchor sliding cache (the
  per-anchor scoring is amortized to O(L) across pivots) but
  recompute the per-window aggregate per pivot (O(W)). Expect
  roughly 10-15x slower than `pwm` at iterator = 1.
- `pwm.grad.ism` is essentially the same cost as `pwm.grad`: the
  three flips contribute O(1) each on top of the per-pivot scan.

For genome-wide attribution, a sparse iterator (peaks, TSS windows)
plus `iterator = 1` *inside* those is usually the right shape. Coarser
iterator steps (`iterator = 10` or `100`) work for whole-genome scans
when single-bp resolution isn't required.
